A community-maintained standard library of population genetic models

### 2019-12-20 popSimManu.tex
some reworking of intro

The \stdpopsim software has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and maintain its quality.
Importantly, we have rigorous quality control methods to ensure implemented models are accurate
and have documented methods for others to contribute new modules.
We invite new collaborators to join our community.
Below we describe the resource and give
examples of how it can be used to benchmark demographic inference methods.

@@

Given the genome data and simulation model descriptions defined within the
library, it is then straightforward to run accurate, standardized simulations
across a range of organisms. \Stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}. Currently,
\stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the default simulation engine. We plan to include other simulation
engines to support scenarios that cannot be modelled under the coalescent framework,
and have already implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} as
an additional backend.

@@

At the time of writing, more models of human history have been implemented than for
any other organism (see Table \ref{tab:catalog}). 
These models include: a simplified version of the \cite{tennessen2012evolution}
model with only the African population specified (expansion from the ancestral
population and recent growth; Africa\_1T12), the three-population model of \cite{gutenkunst2009inferring}
which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
the European and Asian populations (OutOfAfrica\_3G09), the \cite{tennessen2012evolution} two-population variant of the
Gutenkunst model which does not include Asian populations, but more explicitly models
recent rapid human population growth (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
for American populations which specifies ancestral African, European, and Asian population
components (AmericanAdmixture\_4B11), a three-population out-of-Africa model from \cite{ragsdale2019models}
which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19),
a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19),
and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).
Together these models
contain features believed to have widespread impacts in real data (e.g. bottlenecks, population growth,
admixture) and are therefore highly pertinent in the context of method development.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{ragsdale2019models} model and using the
HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

% JK here --- I think it would really help if we described what the models
% were, briefly, rather than saying who inferred them.
In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
run on whole genome simulations under under a sophisticated demographic model
and empirical genetic map and a model of human migration out of Africa
that includes archaic admiture \citep{ragsdale2019models}. In each column of this figure
we show inferred $N(t)$ for the three extant populations in the model.
In each row we show comparisons among the methods (note that we compare two differing
sample sizes for \MSMC).
There is no single ``true'' reference for effective population size
because of model misspecification -- the inference methods are fitting a single population model
to data simulated from mulitple populations.
However, many methods work by matching coalescence time distributions,
and a single-population model with varying population size can match any coalescence time distribution
(in which case coalescence rate is the inverse of the effective size).
For this reason, solid black lines do not show historical census sizes,
but rather inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix).
Blue lines show estimates from each of three replicate whole genome simulations.
While there is variation in accuracy among methods, populations, and individual replicates,
generally the methods are accurate for this model of human history.

\Stdpopsim allows us to readily compare these estimates to those drawn from a different
model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
$N(t)$ from simulations using the same physical and genetic maps, but a different demographic
history that does not include archaic admixture \citep{gutenkunst2009inferring}. Again we see that each
of the methods is capturing relevant parts of the population history, although the
discrepancy between inferred and simulated values varies with $t$. In comparing inferences between the
models it is interesting to note that $N(t)$ estimates for the CHB and CEU
simulated populations are generally better across methods than estimates from the YRI
simulated population.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
\caption{\textbf{Comparing estimates of $N(t)$ in \emph{Drosophila}}. Population
size over time ($N(t)$) estimated from an African population sample. Data were generated by simulating
replicate \emph{D.melanogaster} genomes under the three-epoch \cite{sheehan2016deep} model
with the genetic map inferred in \cite{comeron2012many}. In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_sheehan}
\end{center}
\end{figure}

As most method development for population genetics has been focused on human
data, it is of consequence to ask how such methods might perform in non-human
genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from a demographic
model estimated from an African sample of \emph{D. melanogaster} \citep{sheehan2016deep}.
As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
an empirically derived genetic map, and then try to infer back the parameters of the model.
Accuracy is mixed among methods in this setting and generally worse than what we
observe under simulations of the human genome.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.pdf}
\caption{\textbf{Parameters estimated using a multi-population human model}.
Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp. \textbf{(A)} Data were generated by simulating
replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
inferred in \cite{international2007second}. \textbf{(B)} For \dadi and \fastsimcoal we show parameters inferred
by fitting the depicted IM model, which includes population sizes, migration rates, and a split
time between CEU and YRI samples. \textbf{(C)} Population size estimates for each population (rows)
from \dadi, \fastsimcoal, and \smcpp (columns).
In shades of blue we show $N(t)$ trajectories estimated from each simulation,
and in black the inverse coalescence rates calculated from the true model (see text).
The population split date, $T_{div}$, is shown at
the bottom, with a common X-axis to the population size panels.}
\label{fig:IM_popn_human}
\end{center}
\end{figure}

@@

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the \cite{gutenkunst2009inferring} model. Our results highlight many
of the strengths and weaknesses of the different types of methods we used.
For instance, the SFS-based approaches where we fit simple IM models do not capture
recent exponential growth in the CEU population, but do consistently recover the
simulated YRI population size history. Moreover, these approaches allow for estimating
migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
of divergence times. However, these migration rate estimates are somewhat biased likely
due to model misspecification (Figure \ref{fig:homsap_mig_rates}).
By contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though the inferred population sizes are in general more
noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
estimates that are consistently underestimated (Figure \ref{fig:IM_popn_human}C).

Again, we can compare between species and look at the performance of these methods in
the context of a two-population model from \emph{D. melanogaster}. Figure
\ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
the model of \cite{li2006inferring}, which includes
an ancestral population in Africa, then a population expansion with a population
split and bottleneck into a European population with no post-divergence migration.
Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
parameters, but ignore the brief population bottleneck in Europe. In addition, \dadi
and \fastsimcoal both do reasonably well at correctly inferring the absence of
migration (Figure \ref{fig:dromel_mig_rates}). By contrast, the inferred
demographic parameters from \smcpp are more noisy, though in some cases better
capture the short bottleneck in the Europe population.

@@

We performed demographic analysis for one-and two-population tasks.
Our one population task was to infer historical population size over
time ($N(t)$). This was done using three software packages: \stairwayplot
which uses site frequency spectrum information only \citep{liu2015exploring};
the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
sample sizes $n\in (2 , 8)$; and \smcpp~\citep{terhorst2017robust},
which combines information from the site frequency spectrum with
recombination information as in SMC-based methods. No attempt
was made at trying to optimize the analysis from
any particular software package,
as our goal was not to benchmark performance of methods but
instead show how such benchmarking could be easily done using
the \stdpopsim resource. In this spirit we ran each software package as near
to default parameters as possibile. For \stairwayplot we
set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
``estimate" run mode to infer $N(t)$ with all other parameters set
to their default values. For \MSMC we used the ``--fixedRecombination"
option and used the default number of iterations.

@@

Following simulation, low-recombination portions of chromosomes were masked
from the analysis in a manner that reflects the ``accessible" subset of sites
used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
Specifically we masked regions of 1\,cM or greater in the lowest 5th percentile of the empirical
distribution of recombination, regions which are nearly uniformly absent for
empirical analysis.

@@

From HomSap we ran replicate whole genome simulations under the OutOfAfrica\_3G09 model
with the HapmapII\_GRCh37 genetic map, taking 20 samples each from the Europe and Africa populations.
For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes
with the Comeron2012\_dm6 map due to the large effective population sizes leading to
high effective recombination rates. For this reason, we chose to present only data
from 3 Mb of chromosome 2R for simulations under OutOfAfrica\_2L06.
For the generic IM simulations, we used the HomSap genome along with the
HapmapII\_GRCh37 genetic map and sampled 20 individuals from each population.

@@

\section*{Acknowledgements}
We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible.
Early on in the project we were helped by many people including XXXXXX.
CCK and KEL were funded under NIH Award R35GM119856.
JRA and ADK were funded under NIH Award R01GM117241.
TJS and RNG were funded under NIH Award R01GM127348.
ALG and DRS were funded under NIH award R00HG008696
ND and AS were supported in part by NIH Awards R01HG010346 and R35GM127070.
FR and GG were supported by a Villum Young Investigator award (project no.~00025300).
DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
JK is supported by the Robertson Foundation.

>>>>

The \stdpopsim library has been developed by the PopSim Consortium using a
distributed open source model, with strong procedures in place
to continue its growth and maintain its quality.
Importantly, we have rigorous quality control methods to ensure implemented models are accurate
and have documented methods for others to contribute new modules.
We invite new collaborators to join our community.
Below we describe the resource and give
examples of how it can be used to benchmark demographic inference methods.

@@

Given the genome data and simulation model descriptions defined within the
library, it is then straightforward to run accurate, standardized simulations
across a range of organisms. \stdpopsim has a Python API and a user-friendly
command line interface, allowing users with minimal experience direct access to
state-of-the-art simulations. Simulations are output in the ``tree sequence''
format \citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring}, which
contains complete genealogical information about the simulated samples, is
extremely compact, and can be processed efficiently using the \tskit library
\citep{kelleher2016efficient,kelleher2018efficient}. Currently,
\stdpopsim uses the  \texttt{msprime} coalescent simulator \citep{kelleher2016efficient}
as the default simulation engine. We plan to include other simulation
engines to support scenarios that cannot be modeled under the coalescent framework,
and have already implemented \texttt{SLiM} \citep{haller2019tree,haller2019slim} as
an additional backend.

@@

Currently, more models of human history have been implemented than for
any other organism (see Table \ref{tab:catalog}). 
These models include: a simplified version of the \cite{tennessen2012evolution}
model with only the African population specified (expansion from the ancestral
population and recent growth; Africa\_1T12), the three-population model of \cite{gutenkunst2009inferring}
which specifies the out-of-Africa bottleneck as well as the subsequent divergence of
the European and Asian populations (OutOfAfrica\_3G09), the \cite{tennessen2012evolution} two-population variant of the
Gutenkunst model which does not include Asian populations, but more explicitly models
recent rapid human population growth (OutOfAfrica\_2T12), the \cite{browning2018ancestry} admixture model
for American populations which specifies ancestral African, European, and Asian population
components (AmericanAdmixture\_4B11), a three-population out-of-Africa model from \cite{ragsdale2019models}
which includes archaic admixture (OutOfAfricaArchaicAdmixture\_5R19),
a complex model of ancient Eurasian admixture from \cite{kamm2019efficiently} (AncientEurasia\_9K19),
and a synthetic model of oscillating population size from \cite{schiffels2014inferring} (Zigzag\_1S14).
Together these models
contain features believed to have widespread impacts in real data (e.g. bottlenecks, population growth,
admixture) and are therefore highly pertinent in the context of method development.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairway plot}, and
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the OutOfAfricaArchaicAdmixture\_5R19 model and using the
HapMapII\_GRCh37 genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

% JK here --- I think it would really help if we described what the models
% were, briefly, rather than saying who inferred them.
In Figure \ref{fig:n_t_ragsdale} we present results from our $N(t)$ analysis pipeline
run on whole genome simulations under under OutOfAfricaArchaicAdmixture\_5R19,
a model of human migration out of Africa that includes archaic admixture 
\citep{ragsdale2019models}, along with an empirical genetic map. In each column of this figure
we show inferred $N(t)$ for the three extant populations in the model.
In each row we show comparisons among the methods (note that we compare two differing
sample sizes for \MSMC).
There is no single ``true'' reference for effective population size
because of model misspecification---the inference methods are fitting a single population model
to data simulated from mulitple populations.
However, many methods work by matching coalescence time distributions,
and a single-population model with varying population size can match any coalescence time distribution
(in which case coalescence rate is the inverse of the effective size).
For this reason, solid black lines do not show historical census sizes,
but rather inverse coalescence rates calculated analytically in \texttt{msprime} (see Appendix).
Blue lines show estimates from each of three replicate whole genome simulations.
While there is variation in accuracy among methods, populations, and individual replicates,
generally the methods are accurate for this model of human history.

\stdpopsim allows us to readily compare these estimates to those drawn from a different
model of human history. In Figure \ref{fig:n_t_gutenkunst} we show estimates of
$N(t)$ from simulations using the same physical and genetic maps, but from the OutOfAfrica\_3G09
demographic model that does not include archaic admixture. Again we see that each
of the methods is capturing relevant parts of the population history, although the
discrepancy between inferred and simulated values varies with $t$. In comparing inferences between the
models it is interesting to note that $N(t)$ estimates for the CHB and CEU
simulated populations are generally better across methods than estimates from the YRI
simulated population.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/d_mel_Sheehan_mask2.png}
\caption{\textbf{Comparing estimates of $N(t)$ in \emph{Drosophila}}. Population
size over time ($N(t)$) estimated from an African population sample. Data were generated by simulating
replicate \emph{D.melanogaster} genomes under the African3Epoch\_1S16 model
with the genetic map inferred in \cite{comeron2012many}. In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_sheehan}
\end{center}
\end{figure}

As most method development for population genetics has been focused on human
data, it is of consequence to ask how such methods might perform in non-human
genomes. Figure \ref{fig:n_t_sheehan} shows parameter estimates from the African3Epoch\_1S16
model, originally estimated from an African sample of \emph{D. melanogaster} \citep{sheehan2016deep}.
As with humans, we use \stdpopsim to simulate replicate \emph{Drosophila} genomes under
an empirically derived genetic map, and then try to infer back the parameters of the model.
Accuracy is mixed among methods in this setting and generally worse than what we
observe under simulations of the human genome.

\begin{figure}
\begin{center}
\includegraphics[width=0.7\linewidth]{display_items/homo_sapiens_two_popn_comp.pdf}
\caption{\textbf{Parameters estimated using a multi-population human model}.
Here we show estimates of $N(t)$ inferred using \dadi, \fastsimcoal, or \smcpp. \textbf{(A)} Data were generated by simulating
replicate human genomes under the OutOfAfrica\_3G09 model and using the HapMapII\_GRCh37 genetic map
inferred in \cite{international2007second}. \textbf{(B)} For \dadi and \fastsimcoal we show parameters inferred
by fitting the depicted IM model, which includes population sizes, migration rates, and a split
time between CEU and YRI samples. \textbf{(C)} Population size estimates for each population (rows)
from \dadi, \fastsimcoal, and \smcpp (columns).
In shades of blue we show $N(t)$ trajectories estimated from each simulation,
and in black the inverse coalescence rates calculated from the true model (see text).
The population split date, $T_{div}$, is shown at
the bottom, with a common X-axis to the population size panels.}
\label{fig:IM_popn_human}
\end{center}
\end{figure}

@@

In Figure \ref{fig:IM_popn_human}C we show estimates of population sizes and divergence
time, for each of the inference methods, using samples drawn from African and European populations
simulated under the OutOfAfrica\_3G09 model. Our results highlight many
of the strengths and weaknesses of the different types of methods we used.
For instance, the SFS-based approaches where we fit simple IM models do not capture
recent exponential growth in the CEU population, but do consistently recover the
simulated YRI population size history. Moreover, these approaches allow for estimating
migration rates (Figure \ref{fig:homsap_mig_rates}), also leading to more accurate inference
of divergence times. However, these migration rate estimates are somewhat biased likely
due to model misspecification (Figure \ref{fig:homsap_mig_rates}).
By contrast, \smcpp is much better at capturing the recent exponential
growth in the CEU population, though the inferred population sizes are in general more
noisy. In addition, the assumption of no migration by \smcpp leads to divergence time
estimates that are consistently underestimated (Figure \ref{fig:IM_popn_human}C).

Again, we can compare between species and look at the performance of these methods in
the context of a two-population model from \emph{D. melanogaster}. Figure
\ref{fig:two_popn_fly} shows parameter estimates for simulations drawn from
the OutOfAfrica\_2L06 model, which includes
an ancestral population in Africa, then a population expansion with a population
split and bottleneck into a European population with no post-divergence migration.
Here again, we find that \dadi and \fastsimcoal are more consistent in their inferred
parameters, but ignore the brief population bottleneck in Europe. In addition, \dadi
and \fastsimcoal both do reasonably well at correctly inferring the absence of
migration (Figure \ref{fig:dromel_mig_rates}). By contrast, the inferred
demographic parameters from \smcpp are more noisy, though in some cases better
capture the short bottleneck in the Europe population.

@@

We performed demographic analysis for one-and two-population tasks.
Our one population task was to infer historical population size over
time ($N(t)$). This was done using three software packages: \stairwayplot
which uses site frequency spectrum information only \citep{liu2015exploring};
the SMC based \MSMC~\citep{schiffels2014inferring} which we ran with two different
sample sizes $n\in (2 , 8)$; and \smcpp~\citep{terhorst2017robust},
which combines information from the site frequency spectrum with
recombination information as in SMC-based methods. No attempt
was made at trying to optimize the analysis from
any particular software package,
as our goal was not to benchmark performance of methods but
instead show how such benchmarking could be easily done using
the \stdpopsim resource. In this spirit we ran each software package as near
to default parameters as possible. For \stairwayplot we
set the parameters numRuns=1 and dimFactor=5000. For \smcpp we used the
``estimate" run mode to infer $N(t)$ with all other parameters set
to their default values. For \MSMC we used the ``--fixedRecombination"
option and used the default number of iterations.

@@

Following simulation, low-recombination portions of chromosomes were masked
from the analysis in a manner that reflects the ``accessible" subset of sites
used in empirical population genomic studies \citep[e.g.,][]{danecek20111000,langley2012genomic}.
Specifically we masked regions of \SI{1}{cM} or greater in the lowest 5th percentile of the empirical
distribution of recombination, regions which are nearly uniformly absent for
empirical analysis.

@@

From HomSap we ran replicate whole genome simulations under the OutOfAfrica\_3G09 model
with the HapmapII\_GRCh37 genetic map, taking 20 samples each from the Europe and Africa populations.
For DroMel, runtimes under OutOfAfrica\_2L06 were prohibitively slow when simulating whole genomes
with the Comeron2012\_dm6 map due to the large effective population sizes leading to
high effective recombination rates. For this reason, we chose to present only data
from \SI{3}{\mega\byte} of chromosome 2R for simulations under OutOfAfrica\_2L06.
For the generic IM simulations, we used the HomSap genome along with the
HapmapII\_GRCh37 genetic map and sampled 20 individuals from each population.

@@

\section*{Acknowledgments}
We thank the Probabilistic Modeling in Genomics conference organizers with making this collaboration possible.
Early on in the project we were helped by many people including XXXXXX.
CCK and KEL were funded under NIH Award R35GM119856.
JRA and ADK were funded under NIH Award R01GM117241.
TJS and RNG were funded under NIH Award R01GM127348.
ALG and DRS were funded under NIH award R00HG008696.
ND and AS were supported in part by NIH Awards R01HG010346 and R35GM127070.
FR and GG were supported by a Villum Young Investigator award (project no.~00025300).
DODV is funded by a UC MEXUS-CONACYT Collaborative Grant and a DGAPA-PAPIIT grant (PAPIIT-IA200620).
JK is supported by the Robertson Foundation.
